---
title: "RMINC Parallelism"
author: "Chris Hammill"
date: "October 20, 2016"
output: html_document
---
<div style="line-height: 170%">
The vast majority of problems tackled by RMINC as of 2016 are considered
["embarassingly parallel"](https://en.wikipedia.org/wiki/Embarrassingly_parallel) this means that
the problem can be broken into smaller pieces that can be solved completely independently. Take
for example the standard massively-univariate approach to analyzing neuroimages. Each image
is composed of voxels, and a separate model is computed at each voxel. These models are computed
without any dependency on the models being computed at other voxels. Since there is no
interdependency it is easy to split the voxels up between multiple cores on a computer, or 
even between different computers and combine the results after. 

## A Basic Example 

Imagine for example there are some simple images that are 10 x 10 x 10 voxels. 

```{r}
library(RMINC)
suppressPackageStartupMessages(library(dplyr))

## Pull in some example data
getRMINCTestData("./")

## Read a csv containing the paths to the test files and some covariates
image_frame <- read.csv("rminctestdata/test_data_set.csv", stringsAsFactors = FALSE)

## Take a look at a few sample slices through the first volume volume
image_frame$jacobians_fixed_2[1] %>%
  mincGetVolume %>%
  mincArray %>%
  mincPlotSliceSeries(mfrow = c(2,2), begin = 2, end = -2)
```

Now to see if the values at each voxel depend on some variable, body weight for example, we can fit a linear model. Since 
the models are independent we can divide the problem up. I'll use four cores on my computer, each one will fit 250 of the 1000
models, and then the results can be stuck back together.

```{r}
model <- mincLm(jacobians_fixed_2 ~ Body.Weight, data = image_frame, parallel = c("local", 4))

image_frame$jacobians_fixed_2[1] %>%
  mincGetVolume %>%
  mincArray %>%
  mincPlotSliceSeries(statistic = mincArray(model, "tvalue-Body.Weight")
                      , symmetric = TRUE
                      , low = 2, high = 5
                      , begin = 2, end = -2
                      , legend = "t-value", locator = FALSE
                      , mfrow = c(2,2))
```

Because the job is divided between 4 cores the computation goes much faster. There is some overhead to splitting
and recombining the job, so you won't get a 4x speedup, but it will be faster. The argument `parallel = c("local", 4)`
says to run the jobs on 4 cores of the local machine. 

This system is even more powerful if you have a cluster at your disposal. You can divide the problem in to potentially
100s of peices that can be solved independently.

### Parallel Functions

As of Oct. 20th, 2016, The following functions support the parallel option as used above:

1. mincLm
2. pMincApply
3. mincLmer
4. anatGetAll
5. anatLmer
6. vertexApply
7. vertexLmer

Each of these can run sequentially, locally (multicore), or on a cluster. 

## RMINC on a Cluster

Using RMINC on a cluster is now relatively simple. Thanks to the wonderful 
["BatchJobs"](https://github.com/tudo-r/BatchJobs/blob/master/README.md) package 
RMINC can be configured to run on Torque, SGE, Slurm, ad-hoc SSH clusters and more. We and our collaborators typically 
use Torque and SGE but using these other systems is possible with a little configuration effort. 

### Quick Start

To setup RMINC to run on your cluster you can do one of the following:

  1. Use RMINC's `configureMincParallel` with either  "pbs", "sge", or a custom configuration file.
     Both the PBS (torque) and SGE options will load up default configurations that work well for our
     system. Your mileage may vary.
  2. Use BatchJobs `loadConfig` or `setConfig`, with a custom configuration.
  
For example:

```{r, include=FALSE}
setConfig(
  list(default.resources = 
         list(nodes = 1, vmem = "8G", walltime = "01:00:00", extras = "-l q=defdev.q")))
```

```{r}
## Tell RMINC to use the default SGE configuration
configureMincParallel("sge")

## Quietly fit a mincLm using 6 SGE cores
suppressMessages(  
  qmodel <- mincLm(jacobians_fixed_2 ~ Body.Weight, data = image_frame, parallel = c("sge", 6))
)

##  Check the results look the same as the local version
image_frame$jacobians_fixed_2[1] %>%
  mincGetVolume %>%
  mincArray %>%
  mincPlotSliceSeries(statistic = mincArray(qmodel, "tvalue-Body.Weight")
                      , symmetric = TRUE
                      , low = 2, high = 5
                      , begin = 2, end = -2
                      , legend = "t-value", locator = FALSE
                      , mfrow = c(2,2))

```

As of RMINC 1.4.3 there is no configuration necessary within parallel function calls other than to specify the
number of batches. The argument `parallel = c("sge", 6)` is splitting the job into 6 peices. The `"sge"` portion
is checked to see if it is "local" or "snowfall", otherwise it is ignored, relying instead on BatchJobs configuration
to control execution. This contrasts from earlier versions of RMINC.


### Site Configuration

To make it so you do not need to configure BatchJobs each time you open an R session you can do one of
the following:

  1. Use BatchJobs [configuration files](https://github.com/tudo-r/BatchJobs/wiki/Configuration),
     If you are configuring for your site this will likely involve editing the configuration files
     in the BatchJobs package itself.
  2. Use the Rprofile mechanisms (see `?Startup`) to load the following: 
    ```{r, eval = FALSE}
     setHook(packageEvent("RMINC", "attach"),
           function(...) 
              BatchJobs::loadConfig(system.file("parallel/pbs_BatchJobs.R", package = "RMINC")))
    ```
    
For our setup, I organized a module that sets the `R_PROFILE` environment variable (in addition
to the package libraries) that points to a file containing the above code. This instructs
R to load the pbs configuration file when RMINC is attached.

The default PBS configuration file looks like:

```{r, eval = FALSE}
cluster.functions <- makeClusterFunctionsTorque(system.file("parallel/pbs_script.tmpl", package = "RMINC"))
default.resources <-
  list(nodes = 1,
       vmem = "8G",
       walltime = "01:00:00")
```


It is R code that sets the `cluster.functions` and `default.resources` for BatchJobs. The function 
`makeClusterFunctionsTorque` finds a template file that it fills in whenever a new job is created. The
exampled PBS template looks like:

```
## Job Resource Interface Definition
##
## nodes [integer|character]:  Argument to PBS -l nodes, may be an integer number of nodes, or a character vector
##                             of a colon separated list of node arguments e.g "1:ppn=8" for a scinet node
## walltime   [character(1)]:  Walltime for this job e.g. "01:00:00" for 1 hour
## vmem       [character(1)]:  Memory for each job e.g. "2G" for 2 gigabytes
## extra         [character]:  Extra lines to be added to the job script before calling R  
## 'walltime' and 'memory' settings automatically determine the correct queue, you don't have to
## select the queue yourself.
## Default resources can be set in your .BatchJobs.R by defining the variable
## 'default.resources' as a named list.


#PBS -N <%= job.name %>
## merge standard error and output
#PBS -j oe
## direct streams to our logfile
#PBS -o <%= log.file %>

#PBS -l nodes=<%= resources$nodes %>,walltime=<%= resources$walltime %>,vmem=<%= resources$vmem %>
## remove this line if your cluster does not support arrayjobs
## Disable array jobs: #PBS -t 1-<%= arrayjobs %>
#PBS -V  
## Run R:
## we merge R output with stdout from PBS, which gets then logged via -o option
<% if(!is.null(resources$extra)) cat(resources$extra, sep = "\n") %>
R CMD BATCH --no-save --no-restore "<%= rscript %>" /dev/stdout
```

BatchJobs handles the job.name, log.file, and rscript portions of the template. Our default.resources
configuration provides the nodes, walltime, and vmem.

If you do not use PBS or SGE, you can find example configurations on 
[BatchJobs' github](https://github.com/tudo-r/BatchJobs/tree/master/examples) 

If in the middle of a session you need to change you configuration you can always use
`BatchJobs::setConfig`, `BatchJobs::loadConfig`, and `RMINC::configureMincParallel`. 
For example to ask for twice as much walltime you can run:

```{r, eval = FALSE}
setConfig(
  list(default.resources = 
         list(nodes = 1, vmem = "8G", walltime = "02:00:00")))
```

I know the nested lists are cumbersome, but ideally reconfiguring mid-session should be rare.

It is also important to note that with RMINC 1.4.3 nested parallelism (as is required for SciNet)
has been discontinued. It introduced unnecessary fragility into the code. If enough users request
this be supported it will be re-added to a future release.

```{r, include = FALSE}
unlink("rminctestdata.tar.gz")
unlink("rminctestdata/", recursive = TRUE)
```


## Internals For Intrepid Developers

The basics of parallel system is to dispatch one of three commands, the basic sequential form, the local
form, and the queued form. Within the parallel forms the first task is to divide the problem into pieces.

We'll look at vertexApply as a simple case study

```{r, eval = FALSE}
vertexApply <- function(filenames, fun, ..., mask = NULL, parallel = NULL) 
{
  # Load the data
  vertexData <- vertexTable(filenames)
  
  results <- matrixApply(vertexData, fun, ..., mask = mask, parallel = parallel)
  attr(results, "likeFile") <- filenames[1]
  
  results
}
```

First the filenames are converted into a matrix of values with each column representing a subject and each row representing
a vertex. This is shipped off to `matrixApply` an internal alternative to `apply` that can parallelize both locally and on
a queue.

```{r, eval = FALSE}
matrixApply <- function(mat, fun, ..., mask = NULL, parallel = NULL){
  
  if(!is.null(mask)){
    if(length(mask) == 1 && is.character(mask))
      mask <- readLines(mask)
    
    mask_lgl <- mask > .5
    mat <- mat[mask_lgl,]
  }
  
  fun <- match.fun(fun)
  
  apply_fun <- function(sub_matrix){
    if(!is.matrix(sub_matrix))
      sub_matrix <- matrix(sub_matrix, nrow = 1)
    
    apply(sub_matrix, 1, function(x) fun(x, ...)) %>%
      matrix(ncol = nrow(sub_matrix))
  }
  
  if(is.null(parallel)){
    results <- apply_fun(mat)
  } else {
    n_groups <- as.numeric(parallel[2])
    groups <- split(seq_len(nrow(mat)), groupingVector(nrow(mat), n_groups))
    
    if(parallel[1] == "local") {
      results <- 
        quiet_mclapply(groups, function(group){
          apply_fun(mat[group,])
        }, mc.cores = n_groups) %>%
        Reduce(cbind, ., NULL)
    } else {
      reg <- makeRegistry("matrixApply_registry")
      on.exit( tenacious_remove_registry(reg) )
      
      batchMap(reg, function(group){
        apply_fun(mat[group,])
      }, group = groups)
      
      submitJobs(reg)
      waitForJobs(reg)
      
      results <-
        loadResults(reg, use.names = FALSE) %>%
        Reduce(cbind, ., NULL)
    }
  }
  
  # The apply part (transpose to match output of mincApply)
  results <- t(results)
  
  if(!is.null(mask)){
    results_expanded <- matrix(0, nrow = length(mask_lgl), ncol = ncol(results))
    results_expanded[mask_lgl, ] <- results
    results <- results_expanded
  }
  
  if(ncol(results) == 1) dim(results) <- NULL
  
  results
}
```

We'll break it down section by section:

The first block handles numeric and file name masks. This is used to subset the rows to save computation time 

```{r, eval = FALSE}
  if(!is.null(mask)){
    if(length(mask) == 1 && is.character(mask))
      mask <- readLines(mask)
    
    mask_lgl <- mask > .5
    mat <- mat[mask_lgl,]
  }
```

match.fun ensures that the function passed in is a properly bound function.
```{r, eval = FALSE}
fun <- match.fun(fun)
```

The next block is the engine of `matrixApply`, its a function that applies the user supplied function to each row 

```{r, eval = FALSE}
  apply_fun <- function(sub_matrix){
    if(!is.matrix(sub_matrix))
      sub_matrix <- matrix(sub_matrix, nrow = 1)
    
    apply(sub_matrix, 1, function(x) fun(x, ...)) %>%
      matrix(ncol = nrow(sub_matrix))
  }
```

Now we handle dispatch

```{r, eval = FALSE}
if(is.null(parallel)){
    results <- apply_fun(mat)
  } else {
    n_groups <- as.numeric(parallel[2])
    groups <- split(seq_len(nrow(mat)), groupingVector(nrow(mat), n_groups))
    
    if(parallel[1] == "local") {
      results <- 
        quiet_mclapply(groups, function(group){
          apply_fun(mat[group,])
        }, mc.cores = n_groups) %>%
        Reduce(cbind, ., NULL)
    } else {
      reg <- makeRegistry("matrixApply_registry")
      on.exit( tenacious_remove_registry(reg) )
      
      batchMap(reg, function(group){
        apply_fun(mat[group,])
      }, group = groups)
      
      submitJobs(reg)
      waitForJobs(reg)
      
      results <-
        loadResults(reg, use.names = FALSE) %>%
        Reduce(cbind, ., NULL)
    }
  }
```

If parallel is NULL, we just pass the whole matrix into `apply_fun`, otherwise we handle the two parallel cases.
First the number of parallel jobs the user wants is retrieved. Then the row indices are split into groups using
split and RMINC's internal `groupingVector` function. The split returns a list of indices, one per group.

Then if the first portion of parallel is "local", a muffled version of `parallel::mclapply` is called, split
the job up amongst cores. This creates a list of results which are then merged via `Reduce` and cbind. 

The BatchJobs case follows the same logic, although a few more infrastructure functions are required. A
BatchJobs registry is created with `makeRegistry` and registered for removal when the function ends. 
`tenacious_remove_registry` works hard to ensure all evidence of a failed run is scrubbed from the system.
The triplet of `batchMap`, `submitJobs`, and `waitForJobs` instructs the queue to start running the jobs
and wait patiently for them to finish. The results are then loaded with `loadResults` and reduced as above.

After this there is a little result coercion

```{r, eval = FALSE}
  # The apply part (transpose to match output of mincApply)
  results <- t(results)
  
  if(!is.null(mask)){
    results_expanded <- matrix(0, nrow = length(mask_lgl), ncol = ncol(results))
    results_expanded[mask_lgl, ] <- results
    results <- results_expanded
  }
  
  if(ncol(results) == 1) dim(results) <- NULL
  
  results
```

The results are transposed to give results that resemble mincApply, if a mask was used, the results are inflated
back to their original dimensions, and if the result has only a single column it is returned to a vector form.

Then the results are returned. This bubbles back up to vertexApply which tacks on a likeFile attribute and
returns.

Implementing additional parallel functions will follow this general framework:

1. Create an engine function that does the work
2. Check if `parallel` is NULL, if so dispatch the engine function as is
3. Determine how to split the job up into peices that the engine can run on, create a list or vector
   with the appropriate information
4. If `parallel[1]` is "local" run mclapply or its muffled variant on the engine function and the 
   grouping object
5. Otherwise run the BatchJobs commands above
6. Reduce the results to an appropriate format.

Although it is possible `matrixApply` will work for your problem (potentially with a little transposition).
If not `mincLm` is a little harder to read, but has some examples for how to use a mask to parallelize
jobs.


</div>
